Manuscript for Foodborne-Outbreak Project

### 2023-04-21 code/analysis_code/exploratory_analysis.qmd
more work

The outcome of interest we will explore for the training data is hospitalizations.

### Pathogen Type + Etiology vs Hospitalization
```{r fig.height=8,fig.width=8}
# Pathogen Type + Etiology vs Hospitalization
ggplot(training_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% 
         count() %>% mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Pathogen_Type, y = true_hospitalization, 
           group = Simplified_Etiology, fill = Simplified_Etiology)) +
  geom_col() + 
  labs(x = "Pathogen Type", y = "Hospitalizations", 
       title = "Stacked Bar Chart of Hospitalizations by Pathogen & Etiology") +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold", size = 11, hjust = 0.5))

## Save Graph
image_path <- here("results","images_EDA")
ggsave("pathogen_type_etiology_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Template Setup
```{r}
# Template Graph
theme_set(theme_bw())
theme_update(plot.title = element_text(face = "bold", size = 12))
theme_update(plot.title = element_text(hjust = 0.5))
```

@@

## Hospitalizations
ggplot(training_data_final %>% 
         group_by(Year, Pathogen_Type, Hospitalizations) %>% 
         count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Year, y = true_hospitalization, 
           group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + geom_jitter() + 
  labs(x = "Year", y = "Hospitalizations", 
       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type",
       color = "Pathogen Type") + 
  facet_wrap(~ Pathogen_Type) + 
  theme(legend.position = "none")

## Save Graph
ggsave("time_v_hospitalization_pathogen_type.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Bacterial Etiologies for Outbreaks
```{r}
# General
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
         filter(Pathogen_Type == "Bacteria"), 
       aes(x = Simplified_Etiology)) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

@@

# Salmonella
Salmonella_Hosp <- training_data_final %>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  filter(Simplified_Etiology == "Salmonella") %>%
  filter(`IFSAC Category` != "Multiple") %>%
  filter(`IFSAC Category` != "Other") %>%
  count() %>%
  mutate(total_hospitalizations = n * Hospitalizations) %>%
  arrange(desc(total_hospitalizations))

ggplot(Salmonella_Hosp, aes(x=`IFSAC Category`, y = total_hospitalizations))+
  geom_boxplot() + labs(x = "Total Hospitalizations", y = "Food Type", 
                        title = "Hospitalizations by Food Type for Salmonella") +
  coord_flip()
```
### Hospitalizations by State
```{r}
# States vs Hospitalization
state_v_hospitalization_training <- training_data_final %>% 
  group_by(State) %>% 
  mutate(all_hospitalizations = sum(Hospitalizations)) %>% 
  count(all_hospitalizations) %>% 
  select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))

@@

### Food Type and Hospitalizations
```{r}
# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations) %>% arrange(desc(all_hospitalizations)), 
       aes(x = `IFSAC Category`, y = all_hospitalizations, 
           fill = `IFSAC Category`)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
  coord_flip() + theme(legend.position = "none")

@@

### Locations and Hospitalizations
```{r fig.height=8,fig.width=8}
Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Outbreak Location versus Hospitalizations") +
  coord_flip()

@@

### Bacterial Etiologies
```{r}
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + 
  geom_bar() + 
  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency for Illnesses")+
  coord_flip()

@@

### Food Type and Illnesses
```{r}
ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

### Locations and Illnesses
```{r}
# Locations and Illnesses
ggplot(analysis_data_final %>% group_by(Location) %>% 
         mutate(all_illnesses = sum(Illnesses)) %>%
         count(all_illnesses) %>% arrange(desc(all_illnesses)), aes(x = Location, y = all_illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Location")+
  coord_flip()

>>>>

The outcome of interest we will explore for the training data is hospitalizations.
### Template Setup
```{r}
# Template Graph
theme_set(theme_classic())
theme_update(plot.title = element_text(face = "bold", size = 12))
theme_update(plot.title = element_text(hjust = 0.5))
```

### Pathogen Type + Etiology vs Hospitalization
```{r}
# Pathogen Type vs Hospitalization
ggplot(training_data_final %>% 
         group_by(Pathogen_Type, Hospitalizations) %>% 
         count() %>% mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Pathogen_Type, y = true_hospitalization, 
           )) +
  geom_col() + 
  labs(x = "Pathogen Type", y = "Hospitalizations", 
       title = "Bar Chart of Hospitalizations by Pathogen type") +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold", size = 11, hjust = 0.5))

## Save Graph
image_path <- here("results","images_EDA")
ggsave("pathogen_type_v_hospitalizations.png", path = image_path, width = 3000, height = 2000, units = "px")
```

@@

## Hospitalizations VS year
ggplot(training_data_final %>% 
         group_by(Year, Pathogen_Type, Hospitalizations) %>% 
         count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = Year, y = true_hospitalization, 
           group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + geom_jitter() + 
  labs(x = "Year", y = "Hospitalizations", 
       title = "Scatter Plot of Hospitalizations by Year and Pathogen Type",
       color = "Pathogen Type") + 
  facet_wrap(~ Pathogen_Type) + 
  theme(legend.position = "none")

## Save Graph
ggsave("year_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")

## Hospitalizations VS month
ggplot(training_data_final %>% 
         group_by(Numeric_Month, Hospitalizations) %>% 
         count() %>% 
         mutate(true_hospitalization = n * Hospitalizations), 
       aes(x = as.factor(Numeric_Month), y = true_hospitalization)) +
  geom_boxplot() +
  labs(x = "Month", y = "Hospitalizations", 
       title = "Scatter Plot of Hospitalizations by Month") + 
  theme(legend.position = "none")

## Save Graph
ggsave("month_v_hospitalization.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Bacterial Etiologies for Outbreaks
```{r}
# General
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>%
         filter(Pathogen_Type == "Bacteria")%>%
         group_by(Simplified_Etiology)%>%
         mutate(all_hospitalizations=sum(Hospitalizations)),
       aes(x = fct_reorder(Simplified_Etiology,all_hospitalizations))) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", 
       title = "Bar Chart of Bacteria Frequency") + coord_flip()

@@

#  Top 3 Bacterial Etiology: Salmonella, E. coli, and Clostridium
Top_Hosp <- training_data_final %>%
  select(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  group_by(Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>%
  filter(Simplified_Etiology %in% c("Salmonella","Clostridium","E. coli")) %>%
  filter(`IFSAC Category` != "Multiple") %>%
  filter(`IFSAC Category` != "Other") %>%
  count() %>%
  mutate(total_hospitalizations = n * Hospitalizations)

ggplot(Top_Hosp, aes(x=fct_reorder(`IFSAC Category`, total_hospitalizations),y = total_hospitalizations))+
  geom_boxplot() + labs(x = "Food Type", y = "Total Hospitalizations", 
                        title = "Hospitalizations by Food Type for Top 3 Bacterial Etiology") +
  facet_wrap(vars(Simplified_Etiology))+
  coord_flip()
ggsave("top3_bacterial_etiology_v_food_hospitalization_frequency.png", path = image_path, width = 3000, height = 2000, units = "px")
```

### Hospitalizations by State
```{r}
# States vs Hospitalization
state_v_hospitalization_training <- training_data_final %>% 
  group_by(State) %>% 
  mutate(all_hospitalizations = sum(Hospitalizations)) %>% 
  count(all_hospitalizations) %>% 
  select(State, all_hospitalizations) %>% arrange(desc(all_hospitalizations))

@@

### Food Type and Hospitalizations
```{r}
# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other") %>%
         group_by(`IFSAC Category`) %>% 
         mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations), 
       aes(x = fct_reorder(`IFSAC Category`,all_hospitalizations), y = all_hospitalizations, 
           )) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'") +
  coord_flip() + theme(legend.position = "none")

@@

### Locations and Hospitalizations
```{r}
Location_v_Hospitalization <- training_data_final%>%
  group_by(Location)%>%
  mutate(all_hospitalizations = sum(Hospitalizations)) %>%
         count(all_hospitalizations)%>%
  ggplot(aes(x = fct_reorder(Location,all_hospitalizations), y = all_hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Outbreak Location versus Hospitalizations") +
  coord_flip()

@@

### Bacterial Etiologies
```{r}
ggplot(analysis_data_final %>% 
         group_by(Simplified_Etiology) %>% 
         mutate(Illnesses=sum(Illnesses))%>%
         filter(Pathogen_Type == "Bacteria"), aes(x = fct_reorder(Simplified_Etiology,Illnesses))) + 
  geom_bar() + 
  labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency for Illnesses")+
  coord_flip()

@@

### Food Type and Illnesses
```{r}
analysis_data_final %>% filter(!`IFSAC Category` %in% c("Multiple", "Other"))%>%
  filter(Illnesses>0)%>%
  group_by(`IFSAC Category`)%>%
  mutate(Illnesses=sum(Illnesses))%>%
ggplot(aes(x = fct_reorder(`IFSAC Category`,Illnesses), y = Illnesses)) + 
  geom_col() + 
  scale_y_log10()+
  labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

### Locations and Illnesses
```{r}
# Locations and Illnesses
ggplot(analysis_data_final %>% group_by(Location) %>% 
         mutate(all_illnesses = sum(Illnesses)) %>%
         count(all_illnesses), aes(x = fct_reorder(Location,all_illnesses), y = all_illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Location")+
  coord_flip()

### 2023-04-21 code/analysis_code/statistical_analysis.qmd
more work

### Model Evaluation
```{r}
# Plot
random_forest_illness_tuned %>%
  autoplot()

# Final Specifications
random_forest_illness_tuned_best <- random_forest_illness_tuned %>%
  select_best()
random_forest_illness_final <- random_forest_illness_workflow %>%
  finalize_workflow(random_forest_illness_tuned_best)
random_forest_illness_fit <- random_forest_illness_final %>%
  fit(illness_data)

# Plotting Predicted Illness VS Illnesses
random_forest_illness_predict <- random_forest_illness_fit %>%
  predict(illness_data)

>>>>

### Model Evaluation
```{r}
# Plot
random_forest_illness_tuned %>%
  autoplot()
# Final Specifications
random_forest_illness_tuned_best <- random_forest_illness_tuned %>%
  select_best()
random_forest_illness_final <- random_forest_illness_workflow %>%
  finalize_workflow(random_forest_illness_tuned_best)
random_forest_illness_fit <- random_forest_illness_final %>%
  fit(illness_data)
# Plotting Predicted Illness VS Illnesses
random_forest_illness_predict <- random_forest_illness_fit %>%
  predict(illness_data)

### 2023-04-21 products/manuscript/Manuscript2.qmd
more work

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
format:
  html
bibliography: ../references.bib
csl: ../apa.csl
---

@@

Foodborne pathogens can cause serious gastrointestinal infections. About 600 million cases of foodborne disease are reported worldwide, with almost 1/10 of these cases caused by contaminated food [@he2023]. On average within the United States each year, 1 in 6 Americans will acquire foodborne illness; 128,000 will be hospitalized, and 3,000 will die [@centerforfoodsafetyandappliednutrition2022]. Economic loss associated with food poisoning is estimated to be greater than 50 billion US dollars annually. Restaurants are frequently sites attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020; @sarno2021]. Disease etiologies range from bacteria (ex: *Salmonella* spp. and *E. coli*) to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp. [@cdc2022]. Environmental factors such as heavy metals can also cause foodborne illness.

@@

To verify pathogens of greatest concern, common settings associated with food poisoning, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

From this research project, we would like to answer the following questions:

-   What are the most frequent locations that foodborne outbreaks occur?

-   What pathogen species are most frequently associated with foodborne outbreaks?

-   What type of food is most often seen for foodborne outbreaks?

In other words, immediate outcomes of interest are location, disease etiology, and food category. These will act as predictors for the health outcomes of interest: illnesses and hospitalizations.

Our hypotheses include:

-   Outbreaks occur most often in homes rather than restaurants.

-   The most common pathogen associated with foodborne disease outbreaks is *E. coli*.

-   The food most commonly associated with foodborne outbreaks consists of salad.

@@

Two datasets were adopted for this research. The first dataset--which was used to build statistical models for the study--was uploaded onto *Kaggle* in 2017, with 12 variables and more than 16,950 observations spanning almost two decades from 1998 to 2015 [@cdc2017a]. The second dataset--which was used for machine learning--was retrieved from the CDC's National Outbreak Reporting System (NORS) website. This second dataset covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations).

Linear and Poisson regression models were afterwards constructed using the aforementioned predictors and outcomes from the training data. Data was resampled using the bootstrapping approach 1,000 times for better estimates of accuracy and variability of model coefficients. Random forest and bagging models were also trained to predict health outcomes using the Kaggle dataset for the NORS-CDC data. RMSE was used to compare each of the models. The best model built from the training data was then employed for the test data.

@@

### Training Data

```{r Table of Hospitalizations by State}
first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
  rename(Hospitalizations = all_hospitalizations)
first_table <- flextable(first_table)
first_table <- theme_vanilla(first_table)
first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
first_table <- set_table_properties(first_table, layout = "autofit")
first_table
```

![Figure 1. Hospitalizations by food type from 1998-2015.](images/Food_type_v_hospitalization-01.png)

### Test Data

![Figure 2. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)

![Figure 3. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness-01.png)

![Figure 4. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)

@@

@@

Between simple and multivariate regressions, multivariate models performed better with lower associated RMSEs. For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better than multivariate poisson regression models (RMSE = 59.7, 5.64).

![Figure 5. Predicted illnesses versus actual illness count for multivariate linear regression model.](images/Predicted%20Illnesses%20VS%20Illnesses%20by%20linear%20regression%20with%20all%20predictors.png){fig-align="center"}

@@

@@

Food poisoning in the United States poses a great concern to public health in terms of both morbidity and economic costs. Previous surveillance data points to *Salmonella* bacteria and salads as the main cause and medium for foodborne disease outbreaks, along with the most common setting for such outbreaks being a restaurant. These findings will also give future outbreak investigators a first direction to consider in the field.

# References

>>>>

---
title: "Manuscript for Foodborne-Outbreak Project"
author: "Weifan Wu & Kailin (Kai) Chen"
date: "May 2023"
output: word_document
bibliography: ../references.bib
csl: ../apa.csl
---

@@

Foodborne pathogens typically cause serious gastrointestinal infections.  On average, 1 in 6 Americans acquire foodborne illness each year, leading to 128,000 hospitalizations, and 3,000 mortality [@centerforfoodsafetyandappliednutrition2022].  Economic loss associated with foodborne illness is estimated to be greater than 50 billion US dollars annually.  Moreover, it can be particularly challenging to trace the food sources and exposure settings that are associated with these illnesses .  Nevertheless, foodborne disease outbreak surveillance enables the identification of crucial agents that are commonly involved in outbreaks.  In the United States, CDC conducts surveillance for foodborne disease outbreaks through the Foodborne Disease Outbreak Surveillance System [@gould_surveillance_2013].  The primary responsibility for the identification, investigation, and reporting of foodborne disease outbreaks caused by various enteric microorganism, and chemical/toxic agents lies with state, local, and territorial public health departments, in accordance with established scientific protocols and procedures [@hoffman_capacity_2005].  Previous studies have identified that restaurants are sites frequently attributed to mass-outbreaks, and typical foods associated with these outbreaks include improperly-prepared meats and vegetables [@mun2020;@sarno2021].  The etiologies leading to disease encompass a wide range of microorganisms, ranging from bacteria such as *Salmonella* spp. and *E. coli*, to viruses (norovirus and rotavirus), to parasites (*Cryptosporidium* spp. and *Cyclospora* spp), as well as toxins like marine toxins or metal toxins [@cdc2022].  To verify pathogens of greatest concern, common settings associated with foodborne illness, and foods typically associated with such outbreaks, this study was launched as a research project at the University of Georgia using information gathered over a 17-year period from the Centers for Disease Control and Prevention (CDC).

@@

This research project is designed to answer the following questions, 1). the most frequent locations that foodborne outbreaks occur, 2). the most associated etiology agents, and 3). the food vehicle linked to foodborne outbreaks.  In other words, immediate outcomes of interest are location, disease etiology, and food category.  Furthermore, these were employed as predictors for building a model aimed at forecasting the the health outcomes of interest: illnesses and hospitalizations.
Our research postulates three hypotheses based on empirical evidence, 1). foodborne disease outbreaks exhibit higher incidence rates in home settings compared to restaurant settings, 2). *E. coli* represents the most frequently detected pathogenic agent in cases of foodborne disease outbreaks, and 3). salad is the food item most commonly associated with foodborne disease outbreaks.

@@

Two datasets were adopted for this research. The first dataset, which was used to build statistical models for the study, was uploaded onto Kaggle in 2017 and originated in CDC’s National Outbreak Reporting System (NORS) website, with 12 variables and more than 16,950 observations (CDC, 2017). The second dataset, which was used for evaluating the model performance, was retrieved from the CDC’s NORS website. The second data set covers reports of foodborne and waterborne disease outbreaks, as well as enteric disease outbreaks spread by contact with environmental sources, infected animals or persons, and other means from 2009 to 2020.

@@

After processing procedures, an exploratory data analysis was performed to view general relationships between predictors (time, pathogen type, pathogen etiology, state, and food type) and the health outcomes of interest (illnesses and hospitalizations). We observed certain patterns between preditors and outcomes and selected

### Statistical Analysis and Machine Learning

Linear and Poisson regression models were afterwards constructed using the aforementioned predictors and outcomes from the training data. Specifically, We first assessed the univariate relationships between the main outcome `Simple Etiology` and each health outcome `Illnesses` and `Hospitalizations` using linear regression. Multiple linear and Poisson regression models were also performed with each health outcome. Machine learning approaches including random forest and bagging models were also trained to predict health outcomes using the Kaggle dataset for the NORS-CDC data. For machine learning models, we tuned hyperparameters using randomized grid search. The model with the lowest RMSE among all the models was chosen as the best model from the training data, and subsequently utilized for the test data.

@@

After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set. In the United States, Multi-state outbreaks were the most commonly associated with hospitalizations caused by foodborne illnesses, followed by California, Ohio, and New York. Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2). Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2005, as shown in Table 2.
Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks. Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February.
Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3). Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5). In contrast, *E. coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier. 
Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6).
The testing data set revealed a number of trends that were consistent with those found in the training data set. For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8).
However, some shifts were also observed between the two data sets. While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9).
In addition, the testing data set revealed that *E. coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).

@@

Based on the findings of the exploratory analysis, there are clear associations between time (Month, Year), pathogen type, etiology, location, state, and food type (IFSAC Category) with health outcomes. However, since there is a strong correlation between etiology and pathogen types, the latter was removed. Consequently, our predictive models were constructed using the chosen predictors of time (Month, Year), etiology, location, state, and food type (IFSAC Category).

@@

Between univariate and multivariate regressions, multivariate models performed better with lower associated RMSEs. For both health outcomes, the multivariate linear regression models (RMSE = 58.4, 5.50) constructed performed better with  and than multivariate poisson regression models `r $RMSE = 59.7, 5.64).

@@

In addition to the above information, it's worth noting that the data also revealed some interesting trends related to the seasonality and demographics of foodborne illness outbreaks. For instance, the incidence of outbreaks tends to be highest during the summer months, likely due to factors such as increased outdoor food preparation and consumption of raw or undercooked foods.

In exploratory analysis, slight shift regarding the most common food vehicle that led to most foodborne illnesses (Table 2 and Figure 9) was obtained. This highlighted the dynamic nature of foodborne illness outbreaks, and the need for ongoing monitoring and surveillance to identify emerging trends and adapt food safety practices accordingly. Efforts to improve food safety must be responsive to these changing patterns to prevent and control the spread of pathogens.

@@

Food poisoning in the United States poses a great concern to public health in terms of both morbidity and economic costs. Previous surveillance data points to *Salmonella* bacteria and salads as the main cause and medium for foodborne disease outbreaks, along with the most common setting for such outbreaks being a restaurant. This study allowed for a comprehensive understanding of the potential variables that may impact the occurrence and severity of foodborne disease outbreaks. Identifying key patterns and associations enabled the development of more targeted and effective intervention strategies to mitigate the risks of foodborne illness,and enhance public health outcomes. These findings will also give future outbreak investigators a first direction to consider in the field.

# References
# Appendix: Tables and Plots

![Figure 1. Hospitalizations by month from 1998-2015.](images/month_v_hospitalization.png)

![Figure 2. Common contaminated food linked to hospitalizations from 1998-2015.](images/Food_type_v_hospitalization.png)

![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png)
![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png)
![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png)
![Figure 5. Leading locations that result in hospitalizations ](images/Location_v_hospitalization.png)
![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png)
![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png)
![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png)
![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)

```{r}
 first_table <- readRDS("../../results/state_v_hospitalization_training.rds") %>% 
   rename(Hospitalizations = all_hospitalizations)
first_table <- flextable(first_table) 
first_table <- theme_vanilla(first_table) 
first_table <- set_caption(first_table, caption = as_paragraph(colorize("Table 1. Hospitalizations by state from 1998 to 2015.", color = "black")))
first_table <- set_table_properties(first_table, layout = "autofit")
first_table
```
